Cheminformatics: The Computer Science of Chemical Discovery

### 2011-11-21 rebuttal.tex

\comment{P5, c1, l41: I would dispute the assertion that these methods
  "ignore reception interactions" since the aim is to identify a
  correlation of biological response (e.g. pIC50) with chemical
  structure. The biological response is an explicit measurement of
  receptor interactions on the protein.}
\response{We have updated
  the text to explicitly note that QSAR models do not usually take
  into account \emph{receptor features}. However we note that while
  receptor information is implicit in the $IC_{50}$, the value also
  includes other non-receptor related features such as
  permeability. Furthermore, lack of receptor information in QSAR
  models has been noted as the origin of activity cliffs (Guha and Van
  Drie, \textit{J.~Chem.~Inf.~Model.}, \textbf{2008}, \textit{48},
  1716--1728), thus supporting the statement that traditional QSAR
  models do indeed miss important information on receptor-ligand
  interactions.}

\comment{P5, c2, l22: I would mention naive Bayesian classifiers as well as
this is perhaps the most widely applied method in the field.}
\response{We have included the Na\"{i}ve Bayes as well}

\comment{P5, c2, l42: formal citation for one of Hopfinger?s papers is
  required here.}
\response{The relevant reference has been added}

\comment{P5, c2, l58: this assertion is made with no evidence to back
  it up. Why should multi-target models be more reliable? It is not
  clear to me that they should.}
\response{We have rephrased this statement to be less dogmatic. While
  it is true that for the case of linear models, a multivariate
  multiple regression is in general, equivalent to multiple individual
regression models, one could argue that taking into account the
correlation structure between the multiple y's could lead to improved
models in scenarios where a molecule has affinities for multiple
related targets. We admit that the answer is not clear at this point
and hence suggest that this could a topic for research }

### 2011-11-21 OpenQuestions.tex

\section*{Data mining \& machine learning}
\begin{enumerate}
\item evaluation of similarities in a heterogeneous network (JKW: What are heterogenous networks?).  What is a specific example here?
\item integrate chemical structure information with ontologies.  Specific example problem?
\item Classify molecular descriptors up to equivalence, and dependence on one another.
\item systems-level understanding of small molecules (JKW: which systems? This will help being clearer on the challenge).  What does this mean, and what would a specific challenge problem be?
%\item \emph{Efficient molecule browsing, e.g. on scaffold level}.
%
%Chemical Abstract Services have a molecule browsing tool called SubScape, which allows to browse large-scale
%chemical spaces efficiently. What could be large-scale solutions for doing this within (combined and aligned) 
%public databases.
%
\item \emph{Dynamic similarity search on instant binary vectors}.

@@

\item \emph{Large-scale browsing of molecular property spaces, e.g. on scaffold level, side-effect-level, ...}.

  Certain molecules might have hundreds of biological activities,
  side-effects in humans (SIDER database
  \cite{Kuhn_Campillos_Letunic_Jensen_Bork_2010}), or many other
  properties attached to them. What are large-scale mining and
  visulization options?  How can we mine private and public data
  sources at the very same time?
%
\item \emph{Chemical image/text mining in patents (curation)}.

  There are various tools for doing automatic text mining on chemical
  patents. Still, the overall acceptance rate of chemical text mining
  is improvable, since many medicinal chemists are very concerned
  about the data quality of such efforts.  What could be done to
  improve the mining quality, curate the obtained data, and to provide
  confidence level estimations for each molecule coming from patent
  mining? Do require image2structure and text2structure mining also
  data stores for ensuring a sufficient amount of confidence and data
  quality?  How can patent mining be used to create new drugs faster
  or to speed-up collaboration/licensing discussions?

@@

\section*{Databases \& software engineering}
\begin{enumerate}
\item \emph{Large scale conformational databases}

@@

\item \emph{Map/Reduce in cheminformatics}
  Many cheminformatics tasks apply algorithms over large input files
  or across many molecules (similarity and substructure search,
  docking, filtering). A simple way to parallelize this is to chunk
  the input data and let individual threads/nodes process each chunk. A
  trivial solution is to manually chunk the input and submit a series
  of jobs to a scheduling system. Using the Hadoop ecosystem as an
  example, we can ask various questions:
  \begin{itemize}
  \item Can we employ modern frameworks like Hadoop to support
    embarassingly parallel cases (filtering $10^7$ molecule libraries)?
  \item Can we develop a generalized framework that includes the
    requisite cheminformatics tools that allows users to seamlessly
    distribute jobs over Hadoop and other map reduce systems?
  \item Is the latency involved in Hadoop based datastores (HBase)
    worth the ability to handle massive molecule collections and run
    M/R queries across them? 
  \item Going further, are there cheminformatics problems that can make use
  of the map/reduce paradigm at the algorithmic level?
  \end{itemize}
\end{enumerate}

>>>>

\section*{Data mining \& machine learning}
\begin{enumerate}

\item \emph{Small molecules \& phenotypic data}

Phenotypic (where one takes images of cells and then analyzes them
extract numerical features from the image) screens are increasingly
common. These types of screens lead to very large, very high
dimensional datasets.
\begin{itemize}
\item What methods are suitable to perform feature selection and
  modeling of such data
\item Does phenotype-derived data lead to to ``better'' models for
  small molecules compared to the usual structure-derived data?
\end{itemize}

\item \emph{Inverse QSAR}

(statistical) QSAR is an example of traditional data mining where one
correlates a set of independent variables to a dependent
variable. This lets one predict properties of a new molecule based on
its structural features and those of a training set. The inverse QSAR
problem is such that given a molecule structure representation
(usually a descriptor vector), what are the possible input structures
that satisfy the representation. This can be made more complex, by
asking that the input structures also satisfy aexperimental property
range.
\begin{itemize}
\item What methodologies can be devised to address this problem?
\item How does a given methodology allow us to translate the
  descriptor space to a minimal region of the input space?
\item Are there internal features of a modeling algorithm (say
  hyperplanes in a SVM approach) that lets us simplify the problem?
\end{itemize}

%\item \emph{Efficient molecule browsing, e.g. on scaffold level}.
%
%Chemical Abstract Services have a molecule browsing tool called SubScape, which allows to browse large-scale
%chemical spaces efficiently. What could be large-scale solutions for doing this within (combined and aligned) 
%public databases.
%
\item \emph{Dynamic similarity search on instant binary vectors}.

@@

\item \emph{Chemical image/text mining in patents (curation)}.

  There are various tools for doing automatic text mining on chemical
  patents. Still, the overall acceptance rate of chemical text mining
  is improvable, since many medicinal chemists are very concerned
  about the data quality of such efforts.

  \begin{itemize}
  \item What could be done to improve the mining quality, curate the
    obtained data, and to provide confidence level estimations for
    each molecule coming from patent mining?
  \item Do require image2structure and text2structure mining also data
    stores for ensuring a sufficient amount of confidence and data
    quality?
  \item How can patent mining be used to create new drugs faster or to
    speed-up collaboration/licensing discussions?
 \end{itemize}

@@

\section*{Databases \& software engineering}
\begin{enumerate}
\item \emph{Large scale conformational databases}.

@@

\item \emph{Database indexing schemes for chemical representations}.

Databases of chemical structures are ubiqituous, employing a variety
of RDBMS products. Structure based queries (exact match, substructure,
similarity) are dependent on the chemical representation - most
solutions employ a linear string based form (SMILES, InChI) and
depending on the nature queries to be supported, some are preferred
over others. While one can perform linear scans, indexing is key to
efficient query performance.
\begin{itemize}
\item  Since many queries use binary fingerprints as pre-screens, what
  types of indexing schemes can be designed to support queries on
  binary vectors?
\item If we choose to support 3D searches (shape, pharmacophore) what
  indexing scheme will allow us to perform these types of queries
  efficiently?
\end{itemize}

\item \emph{Map/Reduce in cheminformatics}

  Many cheminformatics tasks apply algorithms over large input files
  or across many molecules (similarity and substructure search,
  docking, filtering). A simple way to parallelize this is to chunk
  the input data and let individual threads/nodes process each chunk. A
  trivial solution is to manually chunk the input and submit a series
  of jobs to a scheduling system. Using the Hadoop ecosystem as an
  example, we can ask various questions:
  \begin{itemize}
  \item Can we employ modern frameworks like Hadoop to support
    embarassingly parallel cases (filtering $10^7$ molecule libraries)?
  \item Can we develop a generalized framework that includes the
    requisite cheminformatics tools that allows users to seamlessly
    distribute jobs over Hadoop and other map reduce systems?
  \item Is the latency involved in Hadoop based datastores (HBase)
    worth the ability to handle massive molecule collections and run
    M/R queries across them? 
  \item Going further, are there cheminformatics problems that can make use
  of the map/reduce paradigm at the algorithmic level?
  \end{itemize}
\end{enumerate}

### 2011-11-21 paper.tex

From the above it becomes clear that the nature of chemical problems
is often a complex mixture of combinatorial and continuous
problems. 
% If we oversimplify the nature of chemical structures and
% their properties, we can not describe the world around us in
% sufficient detail. On the other hand, if we try to encode every detail
% (\emph{e.g.}, all possible 3D structures and their properties, all
% metabolites of a structure in humans in different tissues), the
% complexity of collecting, storing, and mining these data increases
% dramatically. 
For the field of cheminformatics to flourish, we need
closer collaboration between chemists (or, more generally, life
scientists) and computer scientists -- where the former need to be
able to pose their problems in a way relevant for practical
applications; and where the latter are able to devise ways of
capturing, storing, and analyzing chemical data which achieve optimal
balances between the different risks in generalizable encodings and
complexity (space and time). The goals of cheminformatics are to
support better chemical decision making by 1. storing and integrating
data in maintainable ways, 2.  providing sufficient open standards and
tools allowing applications and data to be used across heterogeneous
platforms, and 3. mining the many chemical property spaces in a time-
and space-efficient way. We will discuss all three goals in the
following sections.

@@

InChI or other identity-mapping algorithms allow for exact
searching. However, two other practically relevant algorithms for
scientific discovery based on chemical databases are
\emph{substructure searching} and \emph{similarity searching}, which
are required to generalize from the data points contained in a
particular database to other, related molecules. In substructure
searching, the database is searched for a specified wholly contained
part of the structures in the database, while in similarity searching,
the database is searched for structures similar (in some structure or
property space) to a provided search, or query, structure. Chemical
search packages are often implemented and optimized for a given
database technology, for example, the OrChem package is an open source
chemical search package for the Oracle database \cite{rijnbeek2009}.

@@

The preceding section focuses on the choice of representation of
chemical structures. In the quest for the discovery of novel
therapeutic agents, it is essential to study molecules as they are
situated in a biological context, underlying the processes of life, 
as metabolites, cofactors for enzymatic reactions, hormones, signalling
molecules and environmental xenobiotics. This gives rise to the need
to adequately represent and organise chemical data across biological
databases such as those containing pathways, protein information or 
biological samples. Integration and processing of data from such 
disparate domains underlies ``systems'' level biological research.
This gives rise to some additional challenges that standard chemical
structural representation cannot address: firstly, the representation of
\textit{classes of chemical entities} -- since, in many cases, a class of 
compounds behave in a certain context in a biological system rather than
a single molecule with a fully-specified structure -- and for similar
reasons, chemical concepts such as \textit{groups} (parts of molecules); 
secondly, the need to represent \textit{non-structural} groupings of 
compounds of interest, since compounds may bind to the same or similar 
targets or they may have  similar biological functions such as acting as 
an anti-neoplastic -- preventing the development of tumors. With the 
many different biological databases needing to refer to these different
chemical entities in an organised and standardized 
fashion to enable cross-database integration for a whole-systems 
perspective, \textit{ontologies} and other semantic technologies 
are being successfully employed to support annotation, classification and
semantic cross-domain querying of chemical entities within
biological systems. The most widely used ontology for biologically relevant
chemical entities is ChEBI \cite{chebi2010}, containing around 26,500 
entities in an interlinked semantic graph as of November 2011, and
being actively used for annotation in dozens of biological databases. 
Ontologies are based on logical languages of 
varying levels of expressivity, accompanied by sophisticated reasoning
algorithms. They can be used to construct various sorts of semantic
similarity (e.g. similarity in function, similarity in application) 
which complement traditional structural similarity methods \cite{couto2010}.
An open challenge is integrating graph-based chemical
structure information with the wider logic-based ontology information, which 
will allow combined automated reasoning and querying over the different domains 
encoded in a growing number of interlinked ontologies.% \cite{hastingsowled2010}. 
Enabling automated inferencing in such heterogenous linked systems brings us closer to
achieving a systems level understanding of bioactive chemistry \cite{Oprea:2007fk}.

@@

For researchers interested in the development of new algorithms in the
field, a software library that handles chemical structures and related
data is essential. A wide variety of such chemical toolkits are
available, both commercial (which may offer free academic
licenses) and Open Source.
% [JH comment]
% we should standardize our nomenclature for the phrase 'Open Source' in the
% document. Here it is capitalized, earlier it has been uncapitalized 
% ('open source'), and below it is hyphenated! (open-source)
% [JH comment]

@@

The extent of open-source software in cheminformatics has somewhat
lagged behind related fields such as bioinformatics, partly due to the
existence of a lucrative pharmaceutical market for proprietary
software. However, in recent years there has been a rise in the amount
of open-source cheminformatics software. One contribution to this has
been the Blue Obelisk group~\cite{BlueObelisk2011}
(\url{http://blueobelisk.org}) of chemists, who have worked together
to create interoperable open-source chemistry software (among other
activities). Now there are a number of Open-Source chemical toolkits
available including the Chemistry Development Kit
(CDK)~\cite{steinbeck2003}, Open Babel~\cite{openbabel2011}, RDKit and
Indigo. These toolkits are written in Java or C++ (though toolkits
written in the latter invariably have bindings to a variety of
languages). Such toolkits are used, for example, to read/write
chemical file formats, manipulate chemical structures, measure
similarity of molecules, search for substructures, and generate 2D
diagrams. The underlying algorithms used include graph algorithms
(e.g. maximal common substructure, canonical labelling of graphs),
geometrical methods (e.g. Kabsch alignment) and vector manipulation
(e.g. converting coordinates in various systems, 3D structure
generation). As well as being of direct use to practising
cheminformaticians in their day-to-day work, many applications have
been developed that rely on these toolkits to handle chemical data;
for example, the molecular viewer Avogadro
(\url{http://avogadro.openmolecules.net}) uses Open Babel, while the
molecular workbench Bioclipse~\cite{Bioclipse2} uses the CDK.

The various toolkits have many features in common and at the same time
have certain distinguishing features. For example, the CDK implements
a large collection of molecular descriptors. On the other hand, Open
Babel has robust support for 3D structure generation and optimization
and supports the interconversion of a huge number of chemical file
formats. Table~\ref{tab:features} (in the Appendix) compares features
offered by the open-source toolkits currently available.  Given that
there is ample scope for software engineering (algorithm
implementation, code quality and analysis, building systems and so on), both
the CDK and Open Babel are open to new contributions. Both are driven
via public, active mailing lists and public issue tracking systems.

@@

%
\section{Conclusions}
\label{sec:conclusions}
Cheminformatics, the computer science of chemical discovery, is an
industrial and academic discipline with roots dating back to the 19th
century and flowering in the 1960s with the advent of modern computing
technologies. While many key cheminformatics techniques have been
available in the literature since the 1960's, the bulk of
cheminformatics tools and implementations have been proprietary. In
addition, most data sources of chemical information have been
proprietary (usually with restrictive licensing policies) until
recently. There are many reasons for this, primarily revolving around
issues of profits, competitive intelligence and intellectual property,
which are relevant to the pharmaceutical as well as other chemical
industries. This is in contrast to much of bioinformatics, where the
data and tools have been freely available since the inception of the
field. This disparity could be attributed to the fact that 
the outcomes from cheminformatics methods
and software have a more direct impact on profits, in terms of
identifying lead-like compounds and improving the properties of drug
candidates, while bioinformatics software is employed in
upstream areas such as target identification, 
perhaps not as directly related to
possible profits as an actual small molecule.

%Furthermore, chemical data (structure, activity information etc.) has
%been harder to acquire and when done on large scales can be extremely
%intensive in terms of time and effort, whereas bioinformatics data (sequences for
%example) are much easier to obtain. 
While both fields have theoretical
components, it is possible that the free availability of large amounts
of data in bioinformatics drove the development of publicly
available tools to handle it. In contrast, the proprietary nature of
chemical data would imply that tools to process and analyze that data
were of interest primarily to the owners. As a result, the incentive
to make cheminformatics software public has been low and since the
users of it were primarily industrial, commercial software was the
norm.

@@

As we enter the second decade of the 21st century, the cheminformatics
landscape is rapidly shifting.  There now exist free-of-charge
open-source cheminformatics toolkits, applications and open databases
with tens of millions of compounds and containing valuable
experimental bioactivity data. Modern experimental techniques such as
high throughput sequencing have made the generation of large amounts
of structure-activity data much more accessible.  We believe the
prospects for academic and industrial collaboration between chemists
and computer scientists are bright, and we would like to encourage
more computer scientists to participate in cheminformatics research.

There is an admittedly steep learning curve, as significant domain
knowledge is required to attack many problems in cheminformatics due
to the underlying chemistry of the systems being studied. Indeed, many
issues faced by chemists are often detailed and subtle and do not
admit as clean an abstraction as ``algorithms on a string'', the way
many bioinformatics algorithms can be abstracted into a computer
science framework.  As a result, while one can certainly make
significant contributions to cheminformatics by just considering
structures as graphs, one is limited to somewhat abstract
problems. Keeping in mind that cheminformatics is fundamentally a
practical field that serves to help and advance experimental
chemistry, the key challenges require an understanding of the
underlying chemical systems. 
%One could certainly study for a degree in chemistry. Alternatively one could collaborate with a chemist or cheminformatics practitioner.

Nevertheless, because of the increasing availability of tools and
data, the barrier to entry for non-chemists and non-cheminformaticians
is significantly lower than it was a decade ago. Many questions that
benefit from computer science can now be comprehensively tackled: for
a theorist, graph-theoretic questions of three-dimensional
enumeration; for a database designer, more effective search algorithms
and ontologies; for a practical programmer, expansion of the many
open-source cheminformatics projects.  If more chemists think
algorithmically, and more computer scientists think chemically we will be
much better positioned to take not only ``simple'' chemicals into
account (only one identifier and isomer possible per molecule), but
also their complex relationships, transformations, and combinatorial
challenges.

@@

  \caption{Cheminformatics features provided by open-source toolkits.}

>>>>

From the above it becomes clear that the nature of chemical problems
is often a complex mixture of combinatorial and continuous problems.
For the field of cheminformatics to flourish, we need closer
collaboration between chemists (or, more generally, life scientists)
and computer scientists -- where the former need to be able to pose
their problems in a way relevant for practical applications; and where
the latter are able to devise ways of capturing, storing, and
analyzing chemical data which achieve optimal balances between the
different risks in generalizable encodings and complexity (space and
time). The goals of cheminformatics are to support better chemical
decision making by 1. storing and integrating data in maintainable
ways, 2.  providing sufficient open standards and tools allowing
applications and data to be used across heterogeneous platforms, and
3. mining the many chemical property spaces in a time- and
space-efficient way. We will discuss all three goals in the following
sections.

@@

InChI or other identity-mapping algorithms allow for exact
searching. However, two other practically relevant algorithms for
scientific discovery based on chemical databases are
\emph{substructure searching} and \emph{similarity searching}, which
are required to generalize from the data points contained in a
particular database to other, related molecules. In substructure
searching, the database is searched for a specified wholly contained
part of the structures in the database, while in similarity searching,
the database is searched for structures similar (in some structure or
property space) to a provided search, or query, structure. Chemical
search packages are often implemented and optimized for a given
database technology, for example, the OrChem package is an Open Source
chemical search package for the Oracle database \cite{rijnbeek2009}.

@@

In the quest for the discovery of novel therapeutic agents, it is
essential to study molecules as they are situated in a biological
context, as metabolites, cofactors for enzymatic reactions, hormones
and signalling molecules. Thus we need to adequately represent and
organise chemical data across biological databases such as those
containing pathways, protein information or biological
samples. Integration and processing of data from such disparate
domains underlies ``systems'' level biological research.  This leads
to additional challenges that standard chemical structural
representation cannot address: firstly, the representation of
\textit{classes of chemical entities} -- since, in many cases, a class
of compounds behave in a certain context in a biological system rather
than a single molecule with a fully-specified structure -- and for
similar reasons, chemical concepts such as \textit{groups} (parts of
molecules); secondly, the need to represent \textit{non-structural}
groupings of compounds of interest, since compounds may bind to the
same or similar targets or they may have similar biological functions
such as acting as an anti-neoplastic -- preventing the development of
tumors. With the many different biological databases needing to refer
to these different chemical entities in an organised and standardized
fashion to enable cross-database integration for a whole-systems
perspective, \textit{ontologies} and other semantic technologies are
being successfully employed to support annotation, classification and
semantic cross-domain querying of chemical entities within biological
systems. The most widely used ontology for biologically relevant
chemical entities is ChEBI \cite{chebi2010}, containing around 26,500
entities in an interlinked semantic graph as of November 2011, and
being actively used for annotation in dozens of biological databases.
Ontologies are based on logical languages of varying levels of
expressivity, accompanied by sophisticated reasoning algorithms. They
can be used to construct various sorts of semantic similarity
(e.g. similarity in function, similarity in application) which
complement traditional structural similarity methods \cite{couto2010}.
An open challenge is integrating graph-based chemical structure
information with the wider logic-based ontology information, which
will allow combined automated reasoning and querying over the
different domains encoded in a growing number of interlinked
ontologies. % \cite{hastingsowled2010}.
Enabling automated inferencing in such heterogenous linked systems
brings us closer to achieving a systems level understanding of small
molecule activity \cite{Oprea:2007fk}.

@@

For researchers interested in the development of new algorithms in the
field, a software library that handles chemical structures and related
data is essential. A wide variety of such chemical toolkits are
available, both commercial (which may offer free academic
licenses) and Open Source.

@@

The extent of Open Source software in cheminformatics has somewhat
lagged behind related fields such as bioinformatics, partly due to the
existence of a lucrative pharmaceutical market for proprietary
software. However, in recent years there has been a rise in the amount
of Open Source cheminformatics software. One contribution to this has
been the Blue Obelisk group~\cite{BlueObelisk2011}
(\url{http://blueobelisk.org}) of chemists, who have worked together
to create interoperable Open Source chemistry software (among other
activities). Now there are a number of Open Source chemical toolkits
available including the Chemistry Development Kit
(CDK)~\cite{steinbeck2003}, Open Babel~\cite{openbabel2011}, RDKit and
Indigo. These toolkits are written in Java or C++ (though toolkits
written in the latter invariably have bindings to a variety of
languages). Such toolkits are used, for example, to read/write
chemical file formats, manipulate chemical structures, measure
similarity of molecules, search for substructures, and generate 2D
diagrams. The underlying algorithms used include graph algorithms
(e.g. maximal common substructure, canonical labelling of graphs),
geometrical methods (e.g. Kabsch alignment) and vector manipulation
(e.g. converting coordinates in various systems, 3D structure
generation). As well as being of direct use to practising
cheminformaticians in their day-to-day work, many applications have
been developed that rely on these toolkits to handle chemical data;
for example, the molecular viewer Avogadro
(\url{http://avogadro.openmolecules.net}) uses Open Babel, while the
molecular workbench Bioclipse~\cite{Bioclipse2} uses the CDK.

The various toolkits have many features in common and at the same time
have certain distinguishing features. For example, the CDK implements
a large collection of molecular descriptors. On the other hand, Open
Babel has robust support for 3D structure generation and optimization
and supports the interconversion of a huge number of chemical file
formats. Table~\ref{tab:features} (in the Appendix) compares features
offered by the Open Source toolkits currently available.  Given that
there is ample scope for software engineering (algorithm
implementation, code quality and analysis, building systems and so on), both
the CDK and Open Babel are open to new contributions. Both are driven
via public, active mailing lists and public issue tracking systems.

@@

%
\section{Conclusions}
\label{sec:conclusions}
Cheminformatics, the computer science of chemical discovery, is an
industrial and academic discipline with roots dating back to the 19th
century and flowering in the 1960s with the advent of modern computing
technologies. While many key cheminformatics techniques have been
available in the literature since the 1960's, the bulk of
cheminformatics tools and implementations have been proprietary. In
addition, most data sources of chemical information have been
proprietary (usually with restrictive licensing policies) until
recently. There are many reasons for this, primarily revolving around
issues of profits, competitive intelligence and intellectual property,
which are relevant to the pharmaceutical as well as other chemical
industries. This is in contrast to much of bioinformatics, where the
data and tools have been freely available since the inception of the
field. This disparity could be attributed to the fact that the
outcomes from cheminformatics methods and software have a more direct
impact on profits, in terms of identifying lead-like compounds and
improving the properties of drug candidates, while bioinformatics
software is employed in upstream areas such as target identification,
perhaps not as directly related to possible profits as an actual small
molecule. Furthermore, chemical data (structure, activity information
etc.) has been harder to acquire and when done on large scales can be
extremely intensive in terms of time and effort, whereas
bioinformatics data (e.g., sequences) are much easier to obtain.

While both fields have theoretical components, it is possible that the
free availability of large amounts of data in bioinformatics drove the
development of publicly available tools to handle it. In contrast, the
proprietary nature of chemical data would imply that tools to process
and analyze that data were of interest primarily to the owners. As a
result, the incentive to make cheminformatics software public has been
low and since the users of it were primarily industrial, commercial
software was the norm.

@@

As we enter the second decade of the 21st century, the cheminformatics
landscape is rapidly shifting.  There now exist free-of-charge Open
Source cheminformatics toolkits, applications and open databases with
tens of millions of compounds and containing valuable experimental
bioactivity data. Modern experimental techniques such as high
throughput sequencing have made the generation of large amounts of
structure-activity data much more accessible.  We believe the
prospects for academic and industrial collaboration between chemists
and computer scientists are bright, and encourage more computer
scientists to participate in cheminformatics research.

There is an admittedly steep learning curve, as significant domain
knowledge is required to attack many problems in cheminformatics due
to the underlying chemistry of the systems being studied. Indeed, many
issues faced by chemists do not admit as clean an abstraction as
``algorithms on a string'', the way many bioinformatics algorithms can
be abstracted into a computer science framework.  As a result, while
one can certainly make significant contributions to cheminformatics by
just considering structures as graphs, one is limited to somewhat
abstract problems. Keeping in mind that cheminformatics is
fundamentally a practical field that serves to help and advance
experimental chemistry, the key challenges require an understanding of
the underlying chemical systems.

Nevertheless, because of the increasing availability of tools and
data, the barrier to entry for non-chemists and non-cheminformaticians
is significantly lower than it was a decade ago. Many questions that
benefit from computer science can now be comprehensively tackled: for
a theorist, graph-theoretic questions of three-dimensional
enumeration; for a database designer, more effective search algorithms
and ontologies; for a practical programmer, expansion of the many
Open Source cheminformatics projects.  If more chemists think
algorithmically, and more computer scientists think chemically we will be
much better positioned to take not only ``simple'' chemicals into
account (only one identifier and isomer possible per molecule), but
also their complex relationships, transformations, and combinatorial
challenges.

@@

  \caption{Cheminformatics features provided by Open Source toolkits.}
